\documentclass[10pt]{article}

\usepackage[letterpaper,margin=0.85in,top=0.7in,bottom=0.75in]{geometry}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{titlesec}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{microtype}
\usepackage{float}

\hypersetup{
  colorlinks=true,
  linkcolor=blue,
  citecolor=blue,
  urlcolor=blue,
}

\graphicspath{{../results/}{../results/favorita/}{../results/ablation/}{../results/benchmark/}}

\setlength{\parskip}{0.18em}
\setlength{\parindent}{0pt}
\setlength{\textfloatsep}{5pt plus 1pt minus 1pt}
\setlength{\intextsep}{5pt plus 1pt minus 1pt}
\setlength{\floatsep}{5pt plus 1pt minus 1pt}
\setlength{\abovecaptionskip}{2pt}
\setlength{\belowcaptionskip}{1pt}
\titlespacing*{\section}{0pt}{0.4em}{0.1em}
\titlespacing*{\subsection}{0pt}{0.28em}{0.05em}
\titleformat{\section}{\normalfont\large\bfseries}{\thesection}{0.5em}{}
\titleformat{\subsection}{\normalfont\normalsize\bfseries}{\thesubsection}{0.4em}{}
\setlist{nosep, leftmargin=1.2em, topsep=2pt}
\captionsetup{font=small, labelfont=bf, skip=3pt}

\title{\vspace{-2.4em}\textbf{Re-implementing PatchTST and Applying It to Retail Demand Forecasting}}
\author{Yanlei He, Lijuan Luo \\
\small Cornell University \quad CS 5782 Introduction to Deep Learning \quad Spring 2026}
\date{\vspace{-1.6em}}

\begin{document}
\maketitle
\vspace{-1.4em}

\section{Introduction}

We re-implemented the supervised version of \textbf{PatchTST}~[1], which argues that two architectural choices --- \emph{patching} (treating short subseries as Transformer tokens) and \emph{channel-independence} (sharing one Transformer across channels) --- make Transformers competitive with strong linear baselines for long-horizon forecasting. Our project has two goals: (1) reproduce the paper's main supervised benchmarks (Table~3) and ablation (Table~7); and (2) test whether the same architecture transfers to a practical business problem --- \textbf{retail demand forecasting} on the Corporaci\'on Favorita grocery sales dataset~[3]. Beyond reproducing numbers, we use the implementation as a vehicle to study \emph{which} details actually matter when reproducing a Transformer paper.

\section{Method}

\textbf{Model.} PatchTST splits each input channel of length $L{=}336$ into overlapping patches of length $P{=}16$ with stride $S{=}8$, applies RevIN instance normalization, projects each patch to a $d_{\text{model}}{=}128$ token, adds learnable positional embeddings, and passes the sequence through a 3-layer Transformer encoder ($h{=}16$ heads, $d_{\text{ff}}{=}256$). A flat linear head emits the entire forecast horizon at once. Crucially, channel-independence is implemented by reshaping $(B, L, M)$ to $(B \cdot M, L)$ before the encoder, so the same Transformer weights are reused across all $M$ channels. We additionally implemented DLinear~[2], a vanilla encoder--decoder Transformer baseline, and three ablation variants: \texttt{P+CI} (full PatchTST), \texttt{CI Only} (no patching), \texttt{P Only} (channel-mixing), and \texttt{Original} (vanilla Transformer).

\textbf{Implementation findings.} Cross-checking against the official repository revealed three subtle but important details: (i) PatchTST/42 requires \emph{replication padding} so that the final patch count is 42 rather than 41; (ii) the encoder uses \texttt{BatchNorm1d} rather than \texttt{LayerNorm}, contrary to the typical Transformer convention; and (iii) DLinear uses \emph{shared} weights across channels by default. Fixing these closed an early 5--15\% MSE gap and is reflected in all numbers reported below.

\section{Experimental Setup}

We use ETTh1, ETTm1, and Weather for replication, and Favorita for the retail extension (largest Quito store \#44, 33 product families treated as channels, 2013--2016, 1{,}684 days). Splits are chronological (60/20/20) and standardized using train-split statistics only. Training uses MSE loss, Adam, OneCycleLR, batch size 32, and early stopping (patience 10). We report test MSE and MAE in standardized space, following the paper's protocol.

\section{Results}

\subsection{Benchmark replication (Table 3)}

Across all 12 (dataset, horizon) cells in Table~\ref{tab:bench}, our PatchTST beats the vanilla Transformer by a wide margin and beats or matches DLinear, recovering the paper's qualitative ranking. On Weather, PatchTST wins at every horizon and the absolute MSE is in the same range as the paper. On ETTm1, the ranking holds at every horizon but absolute MSE remains $\sim$0.05--0.08 above the paper. On ETTh1, PatchTST beats DLinear at horizon 96 but is marginally behind ($\Delta\mathrm{MSE} \le 0.017$) at the longer horizons. To check whether the residual ETT gap is just initialization noise, we ran PatchTST with three seeds (2021/2022/2023) on ETTh1 and ETTm1: standard deviation across seeds is only \textbf{0.0006--0.0041 MSE}, an order of magnitude smaller than the gap to the paper (e.g., on ETTh1/720, $\sigma{=}0.004$ vs.\ $\Delta_{\text{paper}}{=}{+}0.214$). The mismatch is therefore systematic, not random; we attribute it to script-level hyperparameter overrides in the official codebase that are not visible in the paper text.

\begin{table}[H]
\centering
\footnotesize
\setlength{\tabcolsep}{4.5pt}
\caption{Benchmark replication of PatchTST Table~3 (test MSE / MAE; lower is better). Bold marks the best result among the three implemented models per row. ``Paper'' lists PatchTST/42 numbers from~[1]; Weather is not reported in that row of Table~3.}
\label{tab:bench}
\begin{tabular}{llcccccccc}
\toprule
& & \multicolumn{2}{c}{\textbf{PatchTST (ours)}} & \multicolumn{2}{c}{\textbf{DLinear (ours)}} & \multicolumn{2}{c}{\textbf{Transformer (ours)}} & \multicolumn{2}{c}{\textbf{Paper PatchTST/42}} \\
\cmidrule(lr){3-4}\cmidrule(lr){5-6}\cmidrule(lr){7-8}\cmidrule(lr){9-10}
Dataset & Horizon & MSE & MAE & MSE & MAE & MSE & MAE & MSE & MAE \\
\midrule
ETTh1   & 96  & \textbf{0.442} & \textbf{0.442} & 0.446 & 0.443 & 0.622 & 0.573 & 0.375 & 0.399 \\
ETTh1   & 192 & 0.492 & 0.475 & \textbf{0.491} & \textbf{0.472} & 0.754 & 0.638 & 0.414 & 0.421 \\
ETTh1   & 336 & 0.548 & 0.511 & \textbf{0.539} & \textbf{0.508} & 0.922 & 0.715 & 0.431 & 0.436 \\
ETTh1   & 720 & 0.659 & 0.581 & \textbf{0.642} & \textbf{0.579} & 1.127 & 0.809 & 0.449 & 0.466 \\
\midrule
ETTm1   & 96  & \textbf{0.346} & \textbf{0.380} & 0.363 & 0.383 & 0.591 & 0.544 & 0.290 & 0.342 \\
ETTm1   & 192 & \textbf{0.396} & 0.412 & 0.409 & \textbf{0.409} & 0.637 & 0.578 & 0.332 & 0.369 \\
ETTm1   & 336 & \textbf{0.438} & \textbf{0.437} & 0.454 & \textbf{0.437} & 0.660 & 0.580 & 0.366 & 0.392 \\
ETTm1   & 720 & \textbf{0.500} & \textbf{0.475} & 0.512 & \textbf{0.473} & 0.757 & 0.629 & 0.420 & 0.424 \\
\midrule
Weather & 96  & \textbf{0.150} & \textbf{0.197} & 0.175 & 0.237 & 0.173 & 0.262 & --- & --- \\
Weather & 192 & \textbf{0.196} & \textbf{0.241} & 0.218 & 0.276 & 0.217 & 0.299 & --- & --- \\
Weather & 336 & \textbf{0.254} & \textbf{0.282} & 0.264 & 0.316 & 0.273 & 0.342 & --- & --- \\
Weather & 720 & \textbf{0.319} & \textbf{0.331} & 0.328 & 0.368 & 0.334 & 0.383 & --- & --- \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Ablation (Table 7): patching vs.\ channel-independence}

Both architectural changes contribute, but their relative importance is dataset-dependent (Table~\ref{tab:ablation}). On ETTh1, removing patching (\texttt{CI Only}) costs only $\sim$0.01 MSE, while removing channel-independence (\texttt{P Only}) costs 0.07--0.14 MSE --- channel-independence is the dominant gain. On Weather, the two contribute roughly equally and the gap to the vanilla Transformer is much smaller, since vanilla self-attention already handles Weather's relatively short patterns reasonably well. The full \texttt{P+CI} model wins in all four cells, supporting the paper's central claim that \emph{both} ingredients matter.

\begin{table}[H]
\centering
\footnotesize
\setlength{\tabcolsep}{6pt}
\caption{Ablation of patching (P) and channel-independence (CI) at horizons 96 and 336 (test MSE).}
\label{tab:ablation}
\begin{tabular}{llcccc}
\toprule
Dataset & Horizon & \textbf{P+CI} (full) & CI Only & P Only & Original (vanilla Tx) \\
\midrule
ETTh1   & 96  & \textbf{0.444} & 0.454 & 0.514 & 0.604 \\
ETTh1   & 336 & \textbf{0.542} & 0.548 & 0.678 & 0.958 \\
Weather & 96  & \textbf{0.149} & 0.174 & 0.163 & 0.172 \\
Weather & 336 & \textbf{0.252} & 0.272 & 0.269 & 0.277 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Retail extension: Favorita grocery sales}

We treat each of the 33 product families as a channel and predict daily total unit sales for the next 7, 14, and 30 days (Table~\ref{tab:favorita}). PatchTST achieves the best MSE at all three horizons, a 27--35\% reduction over DLinear and a 2.7--2.9$\times$ reduction over the vanilla Transformer. The vanilla Transformer overfits aggressively (best epoch 6--7 of 17, with 3--12M parameters), confirming that on small, multi-channel retail data the channel-independent patch design is what makes self-attention usable. Interestingly, 30-day MSE is slightly \emph{lower} than 14-day MSE, which we attribute to the longer window averaging out weekly volatility rather than to a replication bug.

\begin{table}[H]
\centering
\footnotesize
\setlength{\tabcolsep}{8pt}
\caption{Favorita retail forecasting (largest Quito store, 33 product families). Best per row in bold.}
\label{tab:favorita}
\begin{tabular}{lcccccc}
\toprule
& \multicolumn{2}{c}{\textbf{PatchTST}} & \multicolumn{2}{c}{\textbf{DLinear}} & \multicolumn{2}{c}{\textbf{Transformer}} \\
\cmidrule(lr){2-3}\cmidrule(lr){4-5}\cmidrule(lr){6-7}
Horizon & MSE & MAE & MSE & MAE & MSE & MAE \\
\midrule
7-day  & \textbf{0.940} & \textbf{0.566} & 1.302 & 0.671 & 2.736 & 0.997 \\
14-day & \textbf{0.995} & \textbf{0.579} & 1.366 & 0.696 & 2.846 & 1.031 \\
30-day & \textbf{0.930} & \textbf{0.594} & 1.424 & 0.742 & 2.527 & 0.995 \\
\bottomrule
\end{tabular}
\end{table}

\enlargethispage{3\baselineskip}
\section{Discussion}

Three takeaways stand out. First, \textbf{reproduction quality is implementation-bound, not idea-bound}: the paper's qualitative claims replicate cleanly, but matching exact numbers requires hyperparameter and normalization choices that live only in the official scripts. Second, the \textbf{ablation pattern is dataset-dependent}: channel-independence dominates on ETT, while patching matters more on Weather, suggesting these ingredients address different failure modes. Third, the architecture \textbf{transfers cleanly to a real retail problem}: with no domain-specific changes, PatchTST cuts 30-day demand-forecast MSE by $\sim$35\% over DLinear on Favorita. A natural next step is to add exogenous retail variables (promotions, holidays, oil prices), since the current model is purely autoregressive.

\vspace{0.3em}
{\footnotesize
\textbf{References.}\hspace{0.3em}
[1]~Y.~Nie, N.~H.~Nguyen, P.~Sinthong, J.~Kalagnanam. A Time Series is Worth 64 Words: Long-term Forecasting with Transformers. \emph{ICLR 2023}.\hspace{0.4em}
[2]~A.~Zeng, M.~Chen, L.~Zhang, Q.~Xu. Are Transformers Effective for Time Series Forecasting? \emph{AAAI 2023}.\hspace{0.4em}
[3]~Corporaci\'on Favorita Grocery Sales --- Kaggle Store Sales: Time Series Forecasting. \href{https://www.kaggle.com/competitions/store-sales-time-series-forecasting}{kaggle.com/competitions/store-sales-time-series-forecasting}.\par}

\end{document}
